---
title: "Phyloseq_GreenGenes"
author: "Maggi Brisbin"
date: "9/25/2018"
output: 
  html_document:
    toc: true # table of content true
    toc_float: true
    toc_depth: 5  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, fig.path='figures/', fig.width=8, fig.height=5 )
```

Load Packages 
```{r loadLibraries, message=FALSE}
library("phyloseq")
library("ggplot2")
library("tidyr")
library("RColorBrewer")
library(reshape2)
library("vegan")
library(qiime2R)
library(DESeq2)
```

set colors
```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
P <- brewer.pal(12, "Paired")
S2 <- brewer.pal(8, "Set2")
S1 <- brewer.pal(8, "Set1")
D2 <- brewer.pal(8, "Dark2")
colors<- c(P, S2, S1,D2)
colors <- rep(colors, 5)
```

# Load Data

```{r}
phyloseq<-qza_to_phyloseq(features="table.qza", tree = "rooted-tree.qza")
```
load metadata:
```{r}
metatable <- read.table("SampleINFO.txt", header = TRUE)
row.names(metatable) <- metatable[[1]]
metatable<- metatable[,(-1)]

META<- sample_data(metatable)
```


load taxonomy:
```{r}
taxonomy <- read.csv("taxonomy.csv", stringsAsFactors = FALSE, header = TRUE)
names(taxonomy) <- c("row", "tax", "Confidence")
row.names(taxonomy) <-taxonomy[[1]]
taxonomy <- taxonomy[,(-1)]

taxonomy <-  separate(taxonomy, tax, c("D0","D1", "D2", "D3", "D4", "D5", "D6", "D7", "D8", "D9", "D10", "D11", "D12", "D13", "D14"), sep = ";", fill = "right")
taxonomy <- taxonomy[,c(1:7)]
taxmat <- as.matrix(taxonomy)
TAX = tax_table(taxmat)
```


add taxonomy to phyloseq object:
```{r}
ps = merge_phyloseq(phyloseq, TAX, META)

```


# GUT Samples 
subset phyloseq object to include F0 and F1 No Sugar Diet and Control Diet
```{r}
psG <-subset_samples(ps,SampleType == 'Gut' & Generation %in% c('F0', 'F1') & AB== "A" & Treatment %in% c("NSD", "CD"))
```
## Taxonomy Barplot
Transform count data to relative abundance (as %) to normalize for differences in library size.

```{r}
psGra<- transform_sample_counts(psG, function(OTU) 100* OTU/sum(OTU))
```

Plot
```{R}
taxabarplot<-plot_bar(psGra, fill = "D1") + scale_y_continuous(expand = c(0, 0)) + ggtitle("") + scale_fill_manual(values=colors) + theme(legend.title=element_blank()) + geom_bar(aes( fill=D3), stat="identity", position="stack")
taxabarplot + theme(legend.position="none")
```

## Filtering
### prevalence
```{r}
prevdf = apply(X = otu_table(psG),
               MARGIN = ifelse(taxa_are_rows(psG), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(psG),
                    tax_table(psG))

#plyr::ddply(prevdf, "D2", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
```

prevalence plot:
```{r}
prevplot1<-ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps),color=D2)) +
  geom_hline(yintercept = 0.025, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) + 
  theme_bw()+
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~D1) + theme(legend.position="none")

prevplot1
```

*Filter at 2.5% prevalence *
```{r}
prevalenceThreshold = 0.025 * nsamples(psG)

keepTaxa = rownames(prevdf)[(prevdf$Prevalence >= prevalenceThreshold)]
ps2 = prune_taxa(keepTaxa, psG)

#table(tax_table(ps2)[, "D2"], exclude = NULL)
```
```{r}
prevdf = apply(X = otu_table(ps2),
               MARGIN = ifelse(taxa_are_rows(ps2), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps2),
                    tax_table(ps2))

#plyr::ddply(prevdf, "D2", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})

prevplot1<-ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps),color=D2)) +
  geom_hline(yintercept = 0.01, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) + 
  theme_bw()+
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~D1) + theme(legend.position="none")

prevplot1
```
### taxonomic

```{r}
ps0 <- subset_taxa(ps2, !is.na(D1) & !D1 %in% c("", "uncharacterized", "p__"))
```


## Taxonomy Barplot 2
Transform count data to relative abundance (as %) to normalize for differences in library size.

```{r}
ps0ra<- transform_sample_counts(ps0, function(OTU) 100* OTU/sum(OTU))
```

Plot
```{R}
taxabarplot<-plot_bar(ps0ra, fill = "D1") + scale_y_continuous(expand = c(0, 0)) + ggtitle("") + scale_fill_manual(values=colors) + theme(legend.title=element_blank()) + geom_bar(aes( fill=D3), stat="identity", position="stack")
taxabarplot + theme(legend.position="none")
```

## Distance and Ordination 
### Bray-Curtis
Determine the Bray-Curtis distances between samples and perform Principal Coordinate Analysis (PCoA). Plot the PCoA. 
```{r}
ordu = ordinate(ps0ra, "PCoA", "bray")
pbc<-plot_ordination(ps0ra, ordu, color="Treatment", shape = "Generation")+theme_bw() +scale_color_manual(values=P)+ geom_point(size=3)
pbc
```

### Weighted Unifrac

```{r}
ordu2 = ordinate(ps0ra, "PCoA", "unifrac", weighted=TRUE)
pwu<-plot_ordination(ps0ra, ordu2, color="Treatment", shape = "Generation")+theme_bw() +scale_color_manual(values=P)+ geom_point(size=3)
pwu
```

### Unweighted Unifrac

```{r}
ordu3 = ordinate(ps0ra, "PCoA", "unifrac", weighted=FALSE)
puu<-plot_ordination(ps0ra, ordu3, color="Treatment", shape = "Generation")+theme_bw() +scale_color_manual(values=P)+ geom_point(size=3)
puu
```

## Adonis PERMANOVA
###F0
```{r}
F0 <- subset_samples(ps0ra, Generation =="F0")
```
#### Bray-Curtis 
```{r}
set.seed(1)
OTUs <- t(data.frame(otu_table(F0))) #get data frame of OTUs from phyloseq object
meta <- subset(metatable, SampleType == 'Gut' & Generation == 'F0' & Treatment %in% c("NSD", "CD") & AB == "A")
adonis(vegdist(OTUs, method = "bray") ~ Treatment, data = meta)
```

###F1
```{r}
F1 <- subset_samples(ps0ra, Generation =="F1")
```
#### Bray-Curtis 
```{r}
set.seed(1)
OTUs <- t(data.frame(otu_table(F1))) #get data frame of OTUs from phyloseq object
meta <- subset(metatable, SampleType == 'Gut' & Generation == 'F1' & Treatment %in% c("NSD", "CD"))
adonis(vegdist(OTUs, method = "bray") ~ Treatment, data = meta)
```

## Differential Abundance Testing (DESeq2)
###F1
```{r}
F1counts <-  subset_samples(ps0, Generation == "F1")
ddse <- phyloseq_to_deseq2(F1counts, ~Treatment)
```
```{r}
ddse2 <- DESeq(ddse, test="Wald", fitType="parametric")
```
```{r}
res<- results(ddse2,  cooksCutoff = FALSE, contrast=c("Treatment", "NSD", "CD"))
alpha = 0.05
sigtab = res[which(res$padj < alpha), ]
head(sigtab)
```
```{r}
dim(sigtab)
```
Only one SV is significantly differentially abundant ...

Taxonomy?
```{r}
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(F1counts)[rownames(sigtab), ], "matrix"))
sigtab
```

Pseudomonas  (same as last round of analysis, I think?)

# MEDIA Samples 
subset phyloseq object to include F0 and F1 No Sugar Diet and Control Diet
```{r}
psM <-subset_samples(ps,SampleType == 'Media' & Generation %in% c('F0', 'F1') & AB== "A" & Treatment %in% c("NSD", "CD"))
```

Transform count data to relative abundance (as %) to normalize for differences in library size.

```{r}
psMra<- transform_sample_counts(psM, function(OTU) 100* OTU/sum(OTU))
```

Plot
```{R}
taxabarplot<-plot_bar(psMra, fill = "D3") + scale_y_continuous(expand = c(0, 0)) + ggtitle("") + scale_fill_manual(values=colors) + theme(legend.title=element_blank()) + geom_bar(aes( fill=D3), stat="identity", position="stack")
taxabarplot + theme(legend.position="none")
```

## Filtering
### prevalence
```{r}
prevdf = apply(X = otu_table(psM),
               MARGIN = ifelse(taxa_are_rows(psM), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(psM),
                    tax_table(psM))

#plyr::ddply(prevdf, "D2", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
```

prevalence plot:
```{r}
prevplot1<-ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps),color=D2)) +
  geom_hline(yintercept = 0.025, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) + 
  theme_bw()+
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~D1) + theme(legend.position="none")

prevplot1
```
No prevalence filtering for media samples (at least for now) - just remove NA at Phylum level

### taxonomic
```{r}
ps0 <- subset_taxa(psM, !is.na(D1) & !D1 %in% c("", "uncharacterized", "p__"))
```

##DESeq2
```{r}
ddse <- phyloseq_to_deseq2(ps0, ~Treatment)
```


```{r}
dds <- ddse[ rowSums(counts(ddse)) > 1, ]
rld <- rlog(dds, blind=TRUE)
```

```{r}
df <- as.data.frame(assay(rld))
```
add taxonomy 
```{r}
tab = cbind(df, as(tax_table(ps0)[rownames(df), ], "matrix"))
tab <- tab[1:36]
```

```{r}
m <- melt(tab, id=c("D0", "D1", "D2", "D3"))
```

plot D1
```{r}
tabplot<-ggplot(m, aes(x=variable, y=value, color = D1)) + geom_point(size=3, alpha = 0.7) + theme(legend.title=element_blank()) + theme_bw() +
   ggtitle("log2 normalized counts ") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))# +scale_color_manual(values=P)
tabplot
```
plot D2
```{r}
tabplot<-ggplot(m, aes(x=variable, y=value, color = D2)) + geom_point(size=3, alpha = 0.7) + theme(legend.title=element_blank()) + theme_bw() +
   ggtitle("log2 normalized counts ") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))# +scale_color_manual(values=P)
tabplot
```
plot D3
```{r}
tabplot<-ggplot(m, aes(x=variable, y=value, color = D3)) + geom_point(size=3, alpha = 0.7) + theme(legend.title=element_blank()) + theme_bw() +
   ggtitle("log2 normalized counts ") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))# +scale_color_manual(values=P)
tabplot
```

*Get rid of Outliers and plot again* 
```{r}
mNOout <- m[m$value < 30,]
```

plot D1
```{r}
tabplot<-ggplot(mNOout, aes(x=variable, y=value, color = D1)) + geom_point(size=3, alpha = 0.7) + theme(legend.title=element_blank()) + theme_bw() +
   ggtitle("log2 normalized counts ") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))# +scale_color_manual(values=P)
tabplot
```

plot D2
```{r}
tabplot<-ggplot(mNOout, aes(x=variable, y=value, color = D2)) + geom_point(size=3, alpha = 0.7) + theme(legend.title=element_blank()) + theme_bw() +
   ggtitle("log2 normalized counts ") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))# +scale_color_manual(values=P)
tabplot
```

plot D3
```{r}
tabplot<-ggplot(mNOout, aes(x=variable, y=value, color = D3)) + geom_point(size=2, alpha = 0.7) + theme(legend.title=element_blank()) + theme_bw() +
   ggtitle("log2 normalized counts ") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))# +scale_color_manual(values=colors)
tabplot
```


